---
title: "Nartey_ADA final Project"
author: "Cecilia Nartey"
date: '2022-12-07'
output: html_document
---


# This project aims to: 
# i, determine whether there is an association between physical activity and asthma among teenagers in California during 2019 - 2020. 
# ii. dtermine whether sex modifies this association 

To start,
# library already installed packages

```{r}
pacman::p_load(haven, foreign, odds.n.ends, blorr, lmtest, car,  broom, ggplot2, tidyverse, jtools,  naniar, VIM, mice, lattice, table1)

```

```{r}
CHISTeen <- read_sav(file = "teen_2020_spss/TEEN.sav")
```
# Data exploration

## check variable types in dataset

```{r}
str(CHISTeen)
```

# get head of dataframe

```{r}
head(CHISTeen)
```
# Data management
# 1. Convert "Inapplicable(-1) and Not ascertained(-9)" to NA and drop NAs

# 2. Select variables that will be needed and rename them
i. Physical activity
~ TD27_P1 : How many days walked home from school in the past week
Renamed  PhysAct1

~ TD30_P1 : How many days bike/skate home from school in the past week
Renamed PhysAct2

~ TD39: Time doing sitting activities on weekends (TV, computer games, talk with friends) on weekends
Renamed SitActTimeWknds

ii. Asthma
~ TB18: Asthma episode/attack in past 12 mos
Renamed AsthAtt12

The above is a subset of the variable:
~ TB5: Doctor ever told has asthma

Covariates
iii. Age
~ SRAGE_P; renamed Age 

iv. Sex
~ SRSEX; renamed Sex

v. Race
~ RaceCN_P; renamed Race

```{r}
# Converting -1 to NAs
CHISTeen <- CHISTeen %>%
    mutate(across(
        ACMDMT_P1:YRUSM2_P1,
        ~na_if(., -1)
    ))
# Converting -9 to NAs
CHISTeen <- CHISTeen %>%
    mutate(across(
        ACMDMT_P1:YRUSM2_P1,
        ~na_if(., -9)
    ))

```

Next step, rename variables of interest

```{r}
# Renaming variables
## Renaming the variables of interest
CHISTeen <- CHISTeen %>%
  rename (Age = SRAGE_P,
          Sex = SRSEX, 
          Race = RACECNT_P1, 
         PhysAct1 = (TD27_P1), 
          PhysAct2 = (TD30_P1), 
          SitActTimeWknds = TD39, 
          )

CHISTeen <- CHISTeen %>%
  rename (AsthAtt12 = TB18,
         DocEverToldAsth = TB5)

```

Next step, select variables of interest

```{r}
# selecting variables of interest

CHISTeen_VAR <- select(CHISTeen, c('Sex', 'Race', 'Age','PhysAct1', 'PhysAct2','SitActTimeWknds', 'AsthAtt12', 'DocEverToldAsth'))

# checking that heading for correct renaming of variables
head(CHISTeen_VAR)
```
# Complete case analysis exclusions, exclude missing data and/or NAs

```{r}
# Dropping NAs
CHISTeenFin <-CHISTeen_VAR %>%
  drop_na(PhysAct1,PhysAct2,Sex,Race,Age,AsthAtt12)

```

# Check and change classes of some variables and recode some that need recoding

```{r}
# changing to numeric
CHISTeen$PhysAct1 <- as.numeric(CHISTeen$PhysAct1 )
class(CHISTeen$PhysAct1 ) # for "sanity" check

CHISTeen$PhysAct2 <- as.numeric(CHISTeen$PhysAct2 )
class(CHISTeen$PhysAct2) # for "sanity" check

CHISTeenFin$PA_score <-CHISTeenFin$PhysAct1+CHISTeenFin$PhysAct2 
summary(CHISTeenFin$PA_score)

# changing to numeric
CHISTeenFin$SitActTimeWknds <- as.numeric(CHISTeenFin$SitActTimeWknds)
class(CHISTeenFin$SitActTimeWknds) # for "sanity" check

# changing to factor
CHISTeenFin$Sex <- as.factor(CHISTeenFin$Sex)
CHISTeenFin$Race <- as.factor(CHISTeenFin$Race)
CHISTeenFin$Age <- as.factor(CHISTeenFin$Age)
CHISTeenFin$AsthAtt12 <- as.factor(CHISTeenFin$AsthAtt12)

#sanity check
class(CHISTeenFin$Sex)
class(CHISTeenFin$Race)
class(CHISTeenFin$Age)
class(CHISTeenFin$AsthAtt12)

# Recoding some variables
CHISTeenFin <- CHISTeenFin %>%
  mutate(Age = case_when(Age== 12~ 2,
                            Age== 13 ~ 3,
                         Age == 14 ~ 4,
                         Age == 15 ~ 5,
                         Age == 16 ~ 6,
                         Age == 17 ~ 7
                         ))


CHISTeenFin <- CHISTeenFin %>%
  mutate(Race = case_when(Race == 3 ~ 1,
                           Race== 5 ~ 2,
                         Race== 6 ~ 3,
                         Race == 7 ~ 3,
                         ))


# Combining walking and biking/skating home from school as one variable for physical activity

CHISTeenFin$PA_cat <- ifelse(CHISTeenFin$PhysAct1 ==0 & CHISTeenFin$PhysAct2  ==0, 0, 1)


table(CHISTeenFin$PA_cat, CHISTeenFin$AsthAtt12, useNA = 'always') # for "sanity" check

```
 
# Outcome variable (Asthma) -  recode to 1 - "yes", 0 - "no"
 
```{r}
# recoding to 1 - "yes", 0 - "no"
CHISTeenFin <- CHISTeenFin %>%
  mutate(AsthAtt12 = case_when(AsthAtt12== 1~ 1,
                            AsthAtt12== 2 ~ 0
                             ))

table(CHISTeenFin$PA_cat, CHISTeenFin$AsthAtt12, useNA = 'always')
```
 
 
# Descriptives table

```{r}
CHISTeenFin$Race <- 
  factor(CHISTeenFin$Race, levels=c(1,2,3),
         labels=c("Asian", "White", "Other&Multi-Racial"))

  
CHISTeenFin$AsthAtt12<- 
  factor(CHISTeenFin$AsthAtt12, levels=c(0,1),
         labels=c("No Asthmatic Episode", "Had Asthmatic Episode"))

CHISTeenFin$PhysAct1<- 
  factor(CHISTeenFin$PhysAct1, levels=c(0,1,3,5),
         labels=c("0 days", "1-2 days", "3-4 days", "5 or more days"))

CHISTeenFin$PhysAct2 <- 
  factor(CHISTeenFin$PhysAct2 , levels=c(0,1,3),
         labels=c("0 days", "1-2 days" , "3-5 days"))

CHISTeenFin$Sex<- 
  factor(CHISTeenFin$Sex, levels=c(1,2),
         labels=c("Male","Female"))

CHISTeenFin$Age <- 
  factor(CHISTeenFin$Age, levels=c(2,3,4,5,6,7), 
  labels=c("12 year olds", "13 year olds", "14 year olds", "15 year olds", "16 year olds", "17 year olds"))

CHISTeenFin$Race<-setLabel(CHISTeenFin$Race, "Race/Ethnicity")
CHISTeenFin$Sex<-setLabel(CHISTeenFin$Sex, "Sex")
CHISTeenFin$Age<-setLabel(CHISTeenFin$Age, "Age")
CHISTeenFin$AsthAtt12<-setLabel(CHISTeenFin$AsthAtt12, "Asthma Status Episodes")
CHISTeenFin$PhysAct1<-setLabel(CHISTeenFin$PhysAct1, "Days walked home from school")
CHISTeenFin$PhysAct2<-setLabel(CHISTeenFin$PhysAct2, "Days biked/skated home from school")
CHISTeenFin$SitActTimeWknds<-setLabel(CHISTeenFin$SitActTimeWknds, "Amount of inactivity time")

units(CHISTeenFin$Age) <- "years"

Table1 <- table1 (~ Age + Sex + Race + PhysAct1 + PhysAct2 + SitActTimeWknds |AsthAtt12, overall="Total", CHISTeenFin )

```
# Copy table1 to word doc or powerpoint 
```{r}
library(flextable) 
library(magrittr)
t1flex(Table1)%>%
  save_as_docx(path = "table1.docx")
```


# Construct Figure 1 to illustrate exclusion process

```{r}

# install necessary packages

install.packages("DiagrammeR")
library(DiagrammeR) 


grViz("digraph flowchart {

      # node definitions with substituted label text
      node [fontname = Helvetica, shape = rectangle, fontsize=10] 
      # notes indicate how many boxes you will have in your diagram. Since I have two sets of exclusions, I will have three nodes.
      tab1 [label = '@@1'] # starting
      tab2 [label = '@@2'] # exclusion 1
      tab3 [label = '@@3'] # exclusion 2
      
      # edge definitions with the node IDs are used to indicate how the rectangle boxes flow from each other. This is a simple flow chart but I could create branching instead as below
      tab1 -> tab2 -> tab3 
}
      #This set of code provides the text in each rectangle box.
      [1]: 'Records received from CHIS for teenagers in California n=1365'
      [2]: 'Excluding teenagers not diagnosed with asthma n=385'
      [3]: 'Excluding individuals with missing data on variables used in the analysis n=213'
      ") 

```

# Part 3: Data visualization and modeling

## First, let's visually explore the data

```{r}
CHISTeenFin$SitActTimeWknds <- as.numeric(CHISTeenFin$SitActTimeWknds)

library(tidyverse)

# Distribution of asthmatic episodes and amount of inactivity time

ggplot(data = CHISTeenFin) +
  geom_freqpoly(bin=3, aes(x = SitActTimeWknds, color = AsthAtt12)) +
  labs(x = "Inactivity time", title = "Amount of inactivity time on weekends Frequency Polygons by Asthma Status")

```
# Run logistic regression  models

## First, run a univariable (unadjusted) logistic regression for amount of age as a continuous variable

```{r}
# change age back to numeric
CHISTeenFin$Age <- as.numeric(CHISTeenFin$Age)
class(CHISTeenFin$Age)


model1<- glm(AsthAtt12 ~ Age, data=CHISTeenFin, family="binomial")
summary(model1)
```
# Get Odds Ratios
```{r}
odds.n.ends(model1) 
```
# for every one unit increase in age, the odds of experiencing an asthmatic episode increases by 0.87 times. 

# Model sensitivity is 70.4%


# Test linearity assumption
```{r}
CHISTeenFin <- CHISTeenFin %>%
  mutate(Age_log = SitActTimeWknds * log(Age)) # create term to test linearity

model2 <- glm(AsthAtt12 ~ Age + Age_log, data=CHISTeenFin, family="binomial")

summary(model2)

# Since the p-value for the log term (SitActTimeWknds_log) is >.05, we fail to reject the hypothesis that the linearity assumption is not met using the Box Tidwell test, therefore we can use age as a continous or linear variable
```

# Linearity assumption is not violated as the p-value is non-significant (p > 0.05)

# Now lets look at physical activity during the week

```{r}

# first, label levels in PA_cat
CHISTeenFin$PA_cat<- 
  factor(CHISTeenFin$PA_cat, levels=c(0,1),
         labels=c("No days walk or bike/skate from school", "1 or more days walk or bike/skate from scchool"))

baseline_model3 <- glm(AsthAtt12 ~ PA_cat, data=CHISTeenFin, family="binomial")
summary(baseline_model3)
```
# Get odds and ends
```{r}
odds.n.ends(baseline_model3)
```
# Interpretation. OR = 1.24
The odds ratio is 1.24 with a 95% CI of 0.64 and 2.33
The odds of experiencing an asthmatic episode for teenagers who walk or bike/skate home from school 1 or more days during the week is 24% higher compared to those who did not walk or bike/skate home from school any day in the week.


# Adjusting for age, sex 
```{r}
Adjustedmodel1 <- glm(AsthAtt12 ~ PA_cat + Age+ Sex, data=CHISTeenFin, family="binomial")
summary(Adjustedmodel1)

```
# Get odds and ends
```{r}
odds.n.ends(Adjustedmodel1)

```

# Interpretation of Adjusted OR = 1.27
The odds ratio is 1.24 with a 95% CI of 0.66 and 2.41
After adjusting for age and sex, the odds of experiencing an asthmatic episode for teenagers who walk or bike/skate home 1 more days from school during the week is 27% higher compared to those who did not walk or bike/skate home from school any day in the week.

## Look at assumptions for multicollinearity using the vif function
```{r}

vif(Adjustedmodel1)

```


```{r}
vif(Adjustedmodel1)
```

# VIF values are <5, thus there is no multicollinearity

Now, lets look for any influential values using Cook's Distance
```{r}
# Plot Cooks's Distance

plot(Adjustedmodel1, which = 4, id.n = 3, col="red") # which gives Cook's distance calculations (see https://stat.ethz.ch/R-manual/R-patched/library/stats/html/plot.lm.html)
```

# ## Next we will get Cook's D calculations for each obervation and some other stats using the augment function from the Broom package.

```{r}
Adjustedmodel2.data <- augment(Adjustedmodel1) %>%  
  mutate(index = 1:n()) 

head(Adjustedmodel2.data)
```
## Identify and exclude observations with a Cook's Distance greater than a cutoff, use 3 x the mean Cook's D 

```{r}
cuttoff <- mean(Adjustedmodel1.data$.cooksd)*3 # defines cutoff variable

CHISTeenFin_out <- Adjustedmodel1.data %>% # excludes observations
  filter(.cooksd<cuttoff)

nrow(CHISTeenFin_out)/nrow(CHISTeenFin) # check percent kept
# 88.6% of the observations are kept
# many observations are excluded
```

# set cutoff to 0.01

```{r}
cuttoff <- 0.01 # defines cutoff variable

CHISTeenFin_out2 <- Adjustedmodel1.data %>% # excludes observations
  filter(.cooksd<cuttoff)

nrow(CHISTeenFin_out2)/nrow(CHISTeenFin)
```


# Compare coefficients

```{r}
compareCoefs(Adjustedmodel1, Adjustedmodel2)
```
# Now, lets do a model fit assessment using 
```{r}

blr_test_hosmer_lemeshow(Adjustedmodel1)

```
# Non-significant p-value implies that model fitness is good


# Compare the full model to the reduced model with the lrtest function
```{r}
# likelihood ratio test: compare two nested models
lrtest(Adjustedmodel1, Adjustedmodel2)
```


## Let's see how well our model performs for classification and prediction

```{r}
odds.n.ends(Adjustedmodel1, rocPlot = TRUE, predProbPlot = TRUE)
```

# Construct model summary table

```{r}
# First, install packages
install.packages("modelsummary")
install.packages("flextable")

library(modelsummary)
library(flextable)

models = list(baseline_model3, Adjustedmodel1)
modelsummary(models, exponentiate = TRUE, stars = T, statistic = 'conf.int', conf_level = .95)
modelsummary(models, exponentiate = TRUE, stars = T, statistic = 'conf.int', conf_level = .95, footnote='OR = ODDs Ratio', output = "table.docx")

```


# Assess for whether sex modifies the relationship between physical activity and asthma
```{r}

# Change age back to numeric 
CHISTeenFin$Age <- as.numeric(CHISTeenFin$Age)
class(CHISTeenFin$Age)

# Models with and without interaction terms between sex and physical activity 
Adjustedmodel1 <- glm(AsthAtt12 ~ PA_cat + Age+ Sex, data=CHISTeenFin, family="binomial")
summary(Adjustedmodel1)
odds.n.ends(Adjustedmodel1)

model5<- glm(AsthAtt12 ~ PA_cat + Age+ Sex + Sex*PA_cat, data=CHISTeenFin, family="binomial")
summary(model5) 
odds.n.ends(model5)

# Test the hypothesis with the lrtest
lrtest(Adjustedmodel1, model5)
lrtest(model5,Adjustedmodel1)
```

# interaction term via the lr test is not significant so I will not go on to run models stratified by sex. 


